---
title: "Alzheimer's Disease - Data Analysis"
author: "Rishi Dey Chowdhury"
date: "1/2/2022"
output: 
  html_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(GGally)
library(e1071)
library(caret)
library(modelr)
library(yardstick)
library(jtools)
library(knitr)
library(kableExtra)
library(viridis)
options(dplyr.summarise.inform = FALSE)
```

```{r raw_data, cache=TRUE, include=FALSE}
# Importing my data sets
data <- read_csv("AD.csv")
```

# INTRODCTION

In this Data Analysis Project, I am going to work with **Alzheimer's Disease(AD) and Factors affecting it**. The Project will mainly be focused around building a **Logistic Regression Model** to fit to the data.

## MOTIVE

-   To understand how the parameter levels, scores and factors affect a person, in terms of them having AD or not.

## DATA OVERVIEW

Let's take a look at the Data

```{r view_data, warning=FALSE}
kable(head(data), caption = "First few rows of AD Data") %>%
  kable_styling(position = "center")
```

- The Dataset contains `r nrow(data)` rows each representing one individual's measure of `r ncol(data)` parameters.
- The Dataset is obtained from ADNI.
- The Dataset consists of a Healthy Control Group and AD patients.
The description of each column is given below:
```{r column_description}
col.desc <- tibble(
  "column names" = colnames(data),
  "Description" = c("AD Diagnosis Result- 0: AD negative, 1: AD positive", "Age", "Gender", "Education Level", "Biomarker measure after administration of 18F-FDG", "Biomarker measure after administration of AV45", "Nomalized Hippocampus Volume", "APOE2 Gene- 0: Variant Absent, 1: Variant Present", "APOE4 Gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "Mental Medical Examination Score", "Neurobattery Score", "ID of Individual")
)
kable(col.desc, caption = "Column Description") %>%
  kable_styling(position = "center")
```

So, this wraps up the Introduction and Overview of the data, we now understand what we have in the data. We now proceed to visually inspecting the data.

# VISUAL OVERVIEW

Here we will explore the data visually through plots, diagrams, etc. to understand what parameters affect the diagnosis status and to what degree. Our main focus will be to understand graphically or on an upper level **how well each parameter measures segregate between an AD patient and a healthy patient**.
```{r cont_pred_boxplot_data_prep}
conv.to.fac <- colnames(data)[c(1,3, 8:16, 19)]
# Wrangle
data1 <- data
data1[conv.to.fac] <- lapply(data1[conv.to.fac], as.factor)

data.long <- data1 %>%
  select(-c(e2_1, e4_1, rs3818361, rs744373, rs11136000, rs610932, rs3851179, rs3764650, rs3865444, ID, PTGENDER, PTEDUCAT)) %>%
  gather('AGE', 'FDG', 'AV45', 'HippoNV', 'MMSCORE', 'TOTAL13', key = "cont.preds", value = "values")

```

- From the boxplots given below we see that **AV45, FDG, HippoNV, MMSCORE, TOTAL13 visually appear quite significant in separating between the 2 classes i.e. Healthy Individuals(0) and AD Patients(1)**. This gives us an impression that these measures are important in detecting AD. We may expect these parameters to have greater impact in the Logistic Regression Model which we want to build. **AD patients are observed to have higher AV45 measure, lower FDG measure, lower Hippocampus Volume, lower Mental Examination Scores and higher TOTAL13 biomarker measure compared to Healthy Patients**. 
```{r cont_preds_boxplot, fig.asp=1, fig.width=12}
data.long %>%
  ggplot(aes(x= cont.preds, y = values, fill = DX_bl)) +
  geom_boxplot(outlier.shape = NA)+
  facet_wrap(~cont.preds, scales = "free") +
  labs(
    x = "AD Diagnosis Status",
    y = "Values",
    fill = "AD Status"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"), strip.background = element_blank()) +
  theme(legend.position = "bottom") +
  guides(
    fill = guide_legend(
      nrow = 1
    )
  ) + ggtitle("Continuous Parameter Measures")
```

- Now, taking a look at the gene indicators for each individual, we observe that **all the genes are present in lesser proportion of AD patients compared to healthy individuals, except e4_1 gene. There is a significant difference in the proportion of rs11136000 gene**.
```{r gene_count, fig.asp=1, fig.width=12}
gene_count <- data %>%
  gather(e2_1, e4_1, rs3818361, rs744373, rs11136000, rs610932, rs3851179, rs3764650, rs3865444, key = 'gene', value = 'indicator') %>%
  select(DX_bl, gene, indicator)  %>%
  group_by(DX_bl, gene) %>%
  tally(indicator == 1)

gene_count %>%
    mutate(DX_bl = as.factor(DX_bl), n = n/nrow(data)) %>%
    ggplot(aes(gene, n)) +
    geom_bar(aes(fill = DX_bl), position = "dodge", stat = "identity") +
    labs(
        x = "AD Diagnosis Status",
        y = "Presence Prop(gene = 1)",
        fill = "AD Status"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5), panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"), strip.background = element_blank()) +
    theme(legend.position = "bottom") +
    guides(
        fill = guide_legend(
            nrow = 1
        )
    ) + ggtitle("Gene Presence Count")
``` 

- Now, we must also look into the interaction between the continuous columns to be sure that there is no multicollinearity problems in our Model which we will be fitting. And **we see there is not much correlation between the variables, so no multicollinearity issues and some factors like FDG, HippoNV, AV45, MMSCORE, TOTAL13 as we saw above show more or less clear distinction between the 2 class: AD Positive and AD Negative individuals**.
```{r cont_var_pairplot, fig.asp=1, fig.width=12}
data %>%
  ggpairs(columns = c("AGE", "FDG", "AV45", "HippoNV", "MMSCORE", "TOTAL13"), aes(color = as.factor(DX_bl)), progress = FALSE)
```

