---
title: "Alzheimer's Disease - Data Analysis"
author: "Rishi Dey Chowdhury"
date: "1/2/2022"
output: 
  html_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(GGally)
library(e1071)
library(caret)
library(modelr)
library(yardstick)
library(jtools)
library(knitr)
library(kableExtra)
options(dplyr.summarise.inform = FALSE)
```

```{r raw_data, cache=TRUE, include=FALSE}
# Importing my data sets
data <- read_csv("AD.csv")
```

# INTRODCTION

In this Data Analysis Project, I am going to work with **Alzheimer's Disease(AD) and Factors affecting it**. The Project will mainly be focused around building a **Logistic Regression Model** to fit to the data.

## MOTIVE

-   To understand how the parameter levels, scores and factors affect a person, in terms of them having AD or not.

## DATA OVERVIEW

Let's take a look at the Data

```{r view_data, warning=FALSE}
kable(head(data), caption = "First few rows of AD Data") %>%
  kable_styling(position = "center")
```

- The Dataset contains `r nrow(data)` rows each representing one individual's measure of `r ncol(data)` parameters.
- The Dataset is obtained from ADNI.
- The Dataset consists of a Healthy Control Group and AD patients.
The description of each column is given below:
```{r}
col.desc <- tibble(
  "column names" = colnames(data),
  "Description" = c("AD Diagnosis Result- 0: AD negative, 1: AD positive", "Age", "Gender", "Education Level", "Biomarker measure after administration of 18F-FDG", "Biomarker measure after administration of AV45", "Nomalized Hippocampus Volume", "APOE2 Gene- 0: Variant Absent, 1: Variant Present", "APOE4 Gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "specific gene- 0: Variant Absent, 1: Variant Present", "Mental Medical Examination Score", "Neurobattery Score", "ID of Individual")
)
kable(col.desc, caption = "Column Description") %>%
  kable_styling(position = "center")
```

So, this wraps up the Introduction and Overview of the data, we now understand what we have in the data. We now proceed to visually inspecting the data.

